---
title: 'Class 2: Moving averages and ARMA'
author: Andrew Parnell \newline \texttt{andrew.parnell@mu.ie}   \newline \vspace{1cm}
  \newline \includegraphics[width=3cm]{../maynooth_uni_logo.jpg}
  \newline \vspace{1cm}
  \newline PRESS RECORD 
  https://andrewcparnell.github.io/TSDA/
output:
  beamer_presentation:
    includes:
      in_header: ../header.tex
classoption: "aspectratio=169"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'pdf', fig.height = 5)
options(width = 40)
par(mar=c(3,3,2,1), mgp=c(2,.7,0), tck=-.01,las=1)
pkgs = c('R2jags','rjags', 'lubridate', 'tidyverse','forecast')
lapply(pkgs, library, character.only = TRUE)
```

## Learning outcomes

- Recognise and understand the basic theory behind MA(1) and MA(q) models
- Understand the basic ARMA(p,q) formulation
- Know the basics of using the `forecast` package
- Understand the limitations of ARMA forecasting

## Reminder: The most important slide in the course

\Large

Almost all of time series is based on two ideas:

1. Base your future predictions on previous values of the data
1. __Base your future predictions on how wrong you were in your past predictions__


## Reminder: AR models

- An Autoregressive (AR) model works by making the current data point dependent on the previous value, dampened by a parameter
- The usual likelihood used is:
$$y_t \sim N( \alpha + \beta y_{t-1}, \sigma^2)$$
- $\beta$ is usually constrained (naturally via the data) to lie between -1 and 1. Outside that range the process blows up
- The sample PACF is often a good way of diagnosing if an AR model might be appropriate

## Intro to Moving Average Models

- Moving Average (MA) models are similar to AR models but they depend on the previous residual of the series rather than the value itself
- The previous residual is made up of how well we forecasted the last value of the series
- If the previous residual was large (i.e. our forecast was bad) then we want to make a big change to the next prediction
- If the previous residual was small (i.e. our forecast was good) then we might not want to make much of a change

## Moving average models and the ACF/PACF

- Recall that the sample partial autocorrelation function (PACF) can be used to diagnose whether an AR model is appropriate (and also suggest the order $p$)
- For the MA model, it is the sample autocorrelation function (ACF) that helps determine the order of the model
```{r,echo=FALSE,fig.height=3, fig.width = 5}
q = 1 # Order
T = 100
sigma = 1
alpha = 0
set.seed(123)
theta = runif(q)
y = rep(NA,T)
y[1:q] = rnorm(q,0,sigma)
eps = rep(NA,T)
eps[1:q] = y[1:q] - alpha
for(t in (q+1):T) {
  y[t] = rnorm(1, mean = alpha + sum(theta*eps[(t-q):(t-1)]), sd = sigma)
  eps[t] = y[t] - alpha - sum(theta*eps[(t-q):(t-1)])
}
q = 4 # Order
T = 100
sigma = 1
alpha = 0
set.seed(123)
theta = runif(q)
y2 = rep(NA,T)
y2[1:q] = rnorm(q,0,sigma)
eps = rep(NA,T)
eps[1:q] = y[1:q] - alpha
for(t in (q+1):T) {
  y2[t] = rnorm(1, mean = alpha + sum(theta*eps[(t-q):(t-1)]), sd = sigma)
  eps[t] = y2[t] - alpha - sum(theta*eps[(t-q):(t-1)])
}
par(mfrow=c(1,2))
acf(y,main='MA(1)')
acf(y2,main='MA(4)')
```

## Example 1: MA(1)

- The MA(1) model is defined as:
$$y_t = \alpha + \theta \epsilon_{t-1} + \epsilon_t$$
where $\epsilon_t \sim N(0,\sigma^2)$ as usual
- Parameter $\alpha$ represents the overall mean, whilst $\theta$ controls the amount of weight placed on previous residuals
- Like the AR model the values of $\theta$ are not expected to be outside (-1, 1), and negative values can sometimes be physically unrealistic
- The likelihood version of the model is:
$$y_t \sim N(\alpha + \theta \epsilon_{t-1}, \sigma^2)$$

## Simulating from the MA(1) process

Below is some simple code to simulate from an MA(1) process. Note that the first values of `y` and `eps` need to be initialised
```{r, include=FALSE}
set.seed(123)
```
```{r}
T = 100 # Number of observations
sigma = 1 # Residual sd
alpha = 0 # Mean
theta = runif(1) # Choose a positive value
y = eps = rep(NA,T)
y[1] = alpha
eps[1] = 0
for(t in 2:T) {
  y[t] = rnorm(1, mean = alpha + theta * eps[t-1], 
               sd = sigma)
  eps[t] = y[t] - alpha - theta * eps[t-1]
}
```

## Time series plot

```{r,fig.align='center'}
plot(1:T,y,type='l')
```

## Fitting MA(1) models

- We can fit an MA(1) model with the `forecast` package like before
```{r}
Arima(y, order = c(0, 0, 1))
```

## Extending to MA(q)

- As with the AR(p) process we can extend this model to have the current value of $y$ depending on more than one previous residual
- The model becomes an MA(q) model with:
$$y_t \sim N(\alpha + \theta_1 \epsilon_{t-1} + \theta_2 \epsilon_{t-2} + \ldots + \theta_q \epsilon_{t-q}, \sigma^2)$$
- The parameters are as before, except there are now $q$ values of $\theta$. 
- Usually when estimated they will decrease with $q$; the older residuals matter less

## Fitting an MA(q) model

```{r}
Arima(y, order = c(0, 0, 3))
```

- Compare the AIC of this model with the previous MA(1) version

## Forecasting an MA value

- You can create a one step ahead forecast for an MA(1) model by:
$$\hat{y}_{t+1} = \hat{\alpha} + \hat{\theta} \epsilon_t$$

- Forecasts of more than one step ahead will be pretty boring, as every future prediction of $\hat{\epsilon}_t$ will be 0

- Thus MA(q) models are only really informative if you are forecasting $q$ steps ahead

## Combining AR and MA into ARMA

- There is no reason why we have to use just AR or MA on their own
- It's possible to combine them together, for example:
$$y_t = \alpha + \beta y_{t-1} + \theta \epsilon_{t-1} + \epsilon_t$$
This is an _Autoregressive Moving Average_ (ARMA) model
- It's often written as ARMA(p,q) where $p$ is the number of AR terms (here 1) and $q$ the number of MA terms (here also 1)
- ARMA models can deal with a very wide variety of flexible time series behaviour, though they remain stationary
- The likelihood format is:
$$y_t \sim N( \alpha + \beta y_{t-1} + \theta \epsilon_{t-1}, \sigma^2 )$$

## Fitting an ARMA(1, 1) model

```{r}
Arima(y, order = c(1, 0, 1))
```

- Compare again with previous models

## The general ARMA(p, q) framework

- The general equation for an ARMA(p, q) model is:

$$y_t = \alpha + \sum_{i=1}^p \beta_i y_{t-i} + \sum_{j=1}^q \theta_j \epsilon_{t-j} + \epsilon_t$$

- The values of $\beta$ and $\theta$ have to be tightly controlled to get a series that is stationary, though this is only really a problem if we want to simulate the time series

- Occasionally you will run into problems with `Arima` because it doesn't use maximum likelihood (by default) to fit the models. It uses something faster and more approximate instead

## Predicting the future with ARMA

- The `forecast` package contains methods to predict into the future
- First create a model (here ARMA(2, 1))
```{r}
my_model = Arima(y, order = c(2, 0, 1))
```
- ...then forecast...
```{r, fig.height = 4}
plot(forecast(my_model,h = 10))
```

# A real-world example

## Steps in a time series analysis

1. Plot the data and the ACF/PACF
1. Decide if the data look stationary or not. If not, perform a suitable transformation and return to 1
1. Guess at a suitable $p$ and $q$ for an ARMA(p, q) model 
1. Fit the model
1. Try a few models around it by increasing/decreasing $p$ and $q$ and checking the AIC (or others)
1. Check the residuals
1. Forecast into the future



## A real example: wheat data

- Let's follow the steps for the wheat data:
```{r}
wheat = read.csv('../../data/wheat.csv')
plot(wheat$year, wheat$wheat, type = 'l')
```

## ACF and PACF 

```{r, fig.height = 5}
par(mfrow = c(1, 2))
acf(wheat$wheat)
pacf(wheat$wheat)
```

- Suggest starting with AR(1) or AR(3)?

## First model

```{r}
Arima(wheat$wheat, order = c(1, 0, 0))
```

## Next models

- Try AR(2), ARMA(1, 1), and ARMA(2, 1)
```{r}
Arima(wheat$wheat, order = c(2, 0, 0))$aic
Arima(wheat$wheat, order = c(1, 0, 1))$aic
Arima(wheat$wheat, order = c(2, 0, 1))$aic
```

- Best one seems to be ARMA(2, 1). (could also try others)

## Check residuals

- Check the residuals of this model
```{r, fig.height = 4}
my_model_ARMA21 = Arima(wheat$wheat, order = c(2, 0, 1))
qqnorm(my_model_ARMA21$residuals)
qqline(my_model_ARMA21$residuals)
```

## Check residuals' ACF and PACF

```{r, fig.height = 5}
par(mfrow=c(1,2))
acf(my_model_ARMA21$residuals)
pacf(my_model_ARMA21$residuals)
```

## Forecast into the future

```{r}
plot(forecast(my_model_ARMA21,h=20))
```

## What happened to the forecasts here?

- Why did the series diverge rapidly away from what you might have expected?

- The answer is that we have fitted a _stationary model_, i.e. one with constant mean and variance

- The model will just slowly reverts back to that mean over time. The speed at which it reverts will depend on the amount of autocorrelation in the series

- The solution to this lies in better identification of the trend. See the next lecture!

## Summary

- MA(q) models are used to create future forecasts based on the error in the previous forecasts
- ARMA models combine AR and MA ideas together
- The `forecast` package allows us to fit all of these models
- We need to be a bit careful with forecasts that assume stationarity - they will mean-revert