---
title: 'Class 3: Integrated models and ARIMA'
author: Andrew Parnell \newline \texttt{andrew.parnell@mu.ie}   \newline \vspace{1cm}
  \newline \includegraphics[width=3cm]{../maynooth_uni_logo.jpg}
  \newline \vspace{1cm}
  https://andrewcparnell.github.io/IntroTSA/
  \newline PRESS RECORD 
output:
  beamer_presentation:
    includes:
      in_header: ../header.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'pdf', fig.height = 5)
par(mar=c(3,3,2,1), mgp=c(2,.7,0), tck=-.01,las=1)
pkgs = c('R2jags','rjags', 'lubridate', 'tidyverse','forecast', 'rstan')
lapply(pkgs, library, character.only = TRUE)
```

## Learning outcomes

- Understand how differencing works to help make data stationary
- Know the basics of the ARIMA(p, d, q) framework
- Understand how to fit an ARIMA(p, d, q) model in a realistic setting

## Reminder: stationarity

- A time series is said to be weakly stationary if:

    - The mean is stable
    - The variance is stable
    - The autocorrelation doesn't depend on where you are in the series

## Reminder: ARMA models

- Combine the autoregressive and the moving average framework into one

- The general equation for an ARMA(p, q) model is:

$$y_t = \alpha + \sum_{i=1}^p \beta_i y_{t-i} + \sum_{j=1}^q \theta_j \epsilon_{t-j} + \epsilon_t$$

## Combining ARMA with the random walk to produce ARIMA

- There is one other time series model we have already met, that of the random walk:
$$y_t = y_{t-1} + \epsilon_t$$
where $\epsilon_t \sim N(0, \sigma^2)$
- We could re-write this as:
$$y_t - y_{t-1} =  \epsilon_t$$
i.e. the _differences_ are random normally-distributed noise

## Differencing 

- Differencing is a great way of getting rid of a trend
- If $y_t \approx y_{t-1} + b$ then there will be an increasing linear slope in the time series
- Creating $y_t - y_{t-1}$ will remove it and all values will hover around the value $b$
- Even when the trend is non-linear differencing might help
- Differencing twice will remove a quadratic trend for the same reasons
- You can do even higher levels of differencing but this starts to cause problems
- The twice differenced series is:
$$(y_t - y_{t-1}) - (y_{t-1} - y_{t-2}) = y_t - 2 y_{t-1} + y_{t-2}$$

## Idea: combine differencing into the ARMA framework

- We can combine these ideas into the ARMA framework to produce an ARIMA model (the I stands for integrated, i.e. differenced)
- An ARIMA model isn't really stationary as the differences are actually removing part of the trend
- The ARIMA model is written as ARIMA(p,d,q) where $p$ and $q$ are as before and $d$ is the number of differences 

## Example: the ARIMA(1,1,1) model

- If we want to fit an ARIMA(1,1,1) model we first let $z_t = y_t - y_{t-1}$ then fit the model:
$$z_t \sim N( \alpha + \beta z_{t-1} + \theta \epsilon_{t-1}, \sigma^2 )$$
- This is equivalent to an ARMA model on the first differences
- Note that by default `forecast` does not include the term $\alpha$ in the model. You need to add `include.drift = TRUE`

## Fitting an ARIMA(1, 1, 1) model to the wheat data

- Recall that the ARMA(2,1) fit wasn't very good to the wheat data
- Instead try an ARIMA(1, 1, 0) model (i.e. AR(1) on the first differences)

```{r}
wheat = read.csv('../../data/wheat.csv')
Arima(wheat$wheat, order = c(1, 1, 0), 
      include.drift = TRUE)
```

- Much lower AIC!

## General format: the ARIMA(p,d,q) model

- First take the $d$th difference of the series $y_t$, and call this $z_t$
- If you want to do this by hand in R you can use the `diff` function, e.g. `diff(y, differences = 2)` 
- Then fit the model:
$$z_t \sim N \left( \alpha + \sum_{i=1}^p \beta_i z_{t-i} + \sum_{j=1}^q \theta_j \epsilon_{t-j}, \sigma^2 \right)$$

## Choosing $p$, $d$ and $q$

- It's much harder to have an initial guess at all of $p$, $d$ and $q$ in one go 
- We can usually guess at the number of differences $d$ from the time series and ACF plots. If there is a very high degree of autocorrelation it's usually a good idea to try a model with $d$=1 or 2
- I've never met a model where you needed to difference more than twice. Beware of over-differencing

# Revisiting the real-world example

## Steps in an ARIMA time series analysis

1. Plot the data and the ACF/PACF
1. Decide if the data look stationary or not. If not, perform a suitable transformation and return to 1. __If the data has a strong trend or there is a high degree of autocorrelation try 1 or 2 differences__
1. Guess at values of $p$, $d$, and $q$ for an ARIMA(p, d, q) model 
1. Fit the model
1. Try a few models around it by increasing/decreasing $p$, $d$ and $q$ and checking the AIC (or others)
1. Check the residuals
1. Forecast into the future

## A real example: wheat data

- Plot reminder 
```{r}
wheat = read.csv('../../data/wheat.csv')
plot(wheat$year, wheat$wheat, type = 'l')
```

## ACF and PACF 

```{r, fig.height = 5}
par(mfrow = c(1, 2))
acf(wheat$wheat)
pacf(wheat$wheat)
```

- Suggest looking at first differences

## Plot of first differences

```{r}
plot(diff(wheat$wheat), type = 'l')
```

## ACF/PACF of first differences

```{r, fig.height = 5}
par(mfrow = c(1, 2))
acf(diff(wheat$wheat))
pacf(diff(wheat$wheat))
```

- Interesting peaks in ACF at lag 2, and PACF at lag 2. 

## First model

```{r}
Arima(wheat$wheat, order = c(0, 1, 0),
      include.drift = TRUE)
```

- This is just a random walk model. Can also get these from `forecast` with the function `naive`

## Next models
\small
- Try ARIMA(1, 1, 1), ARIMA(1, 1, 0), ARIMA(0, 1, 1)

```{r}
Arima(wheat$wheat, order = c(1, 1, 1), 
      include.drift = TRUE)$aic
Arima(wheat$wheat, order = c(1, 1, 0), 
      include.drift = TRUE)$aic
Arima(wheat$wheat, order = c(0, 1, 1), 
      include.drift = TRUE)$aic
```

- Best one seems to be ARIMA(1, 1, 1). (though BIC suggests others)

## Check residuals

- Check the residuals of this model
```{r, fig.height = 5}
my_model_ARIMA111 = Arima(wheat$wheat, order = c(1, 1, 1),
                          include.drift = TRUE)
qqnorm(my_model_ARIMA111$residuals)
qqline(my_model_ARIMA111$residuals)
```

## Forecast into the future

```{r}
plot(forecast(my_model_ARIMA111, h = 20))
```

## Why do we need to the drift term?

- Without the drift term the forecast will stabilise at or near the first few values of the series
- The MA part of the model is obviously flat (as previously discussed) because there are no further errors to correct
- The AR part of the model reverts back to the estimated mean of the last data point because the $\beta$ parameter is less than 1 - it dampens out the future predictions and stops them from going crazy
- The drift keeps the values going up into the future
- `forecast` doesn't seem to like including the drift/mean when there are multiple differences and AR terms too (not sure why)

## Summary

- ARIMA models extend the ARMA framework to further add in differencing
- ARIMA models are no longer stationary as soon as $d>0$
- A single difference will remove a linear trends, a second difference quadratic trends
- Can spot the need for differencing from the time series plot and the ACF
- Do not over-difference your data!

