---
title: 'Class 3: Linear Regression and GLMs'
author: Andrew Parnell \newline \texttt{andrew.parnell@mu.ie}   \newline \vspace{1cm}
  \newline \includegraphics[width=3cm]{../maynooth_uni_logo.jpg}
  \newline \vspace{1cm}
  https://andrewcparnell.github.io/TSDA/
  \newline PRESS RECORD 
output:
  beamer_presentation:
    includes:
      in_header: ../header.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'pdf', fig.height = 5)
options(width = 40)
par(mar=c(3,3,2,1), mgp=c(2,.7,0), tck=-.01,las=1)
pkgs = c('R2jags','rjags', 'lubridate', 'tidyverse','forecast')
lapply(pkgs, library, character.only = TRUE)
```

## Learning outcomes

- Know how to fit and interpret a linear regression
- Know the difference between a linear regression model and a generalised linear model (GLM)
- Know what a link function is and why it is used
- Be able to interpret the output of a simple GLM

## Using more of the data

- It's very rare that we are given a data set with just a single variable
- More often we're given multiple variables and asked to predict one or more of the variables from the others
- This is an example of _conditional_ inference
- It might look more complicated, but this is still just fitting a probability distribution to some data

## Linear regression example

Here is some data on sea level rise:
```{r}
sl = read.csv('../../data/tide_gauge.csv')
with(sl, plot(year_AD, sea_level_m))
```

## Linear regression models

- The simplest version of a linear regression model has:

    - A _response variable_ ($y$) which is what we are trying to predict/understand
    - An _explanatory variable_ or _covariate_ ($x$) which is what we are trying to predict the response variable from
    - Some _residual uncertainty_ ($\epsilon$) which is the leftover uncertainty that is not accounted for by the explanatory variable
  
- Our goal is to predict the response variable from the explanatory variable, _or_ to try and discover if the explanatory variable _causes_ some kind of change in the response 

## The linear models in maths

- We write the linear model as:
$$y_i = \alpha + \beta x_i + \epsilon_i$$
where $\alpha$ is the intercept, $\beta$ the slope, and $i=1, \ldots, N$ represents each of the $N$ observations

- Usually we make the additional assumption that $\epsilon_i \sim N(0, \sigma^2)$ where $\sigma$ is the residual standard deviation

- Writing this in probability distributions: 
$$y_i|x_i, \alpha, \beta, \sigma \sim N(\alpha + \beta x_i, \sigma^2)$$. 

## Fiting linear regression models

- We can create a likelihood as before by guessing some values of the parameters and then using the `dnorm` function to compute the likelihood value
```{r}
alpha = 2
beta = 1.5
sigma = 0.6
y = sl$sea_level_m
x = sl$year_AD
sum(dnorm(y, mean = alpha + beta*x, sd = sigma, log = TRUE))
```

- Not a very high value of the likelihood!

## Finding the best values

- Luckily, R has the `lm` function to find the best fitting values of the parameters

\tiny
```{r}
summary(lm(y ~ x))
```

## Checking the likelihood

```{r}
alpha = coefficients(lm(y ~ x))[1]
beta = coefficients(lm(y ~ x))[2]
sigma = summary(lm(y ~ x))$sigma
sum(dnorm(y, mean = alpha + beta*x, sd = sigma, 
          log = TRUE))
```

A much higher value of the likelihood!

## Other notes about `lm`

- Usually we would store the output from `lm` in another object to allow us to manipulate the output, e.g. `my_model = lm(y ~ x)`
- We can use the `confint` function to get confidence intervals on the parameters
- We can predict future values of sea level from the model by giving it new $x$ values, e.g. 
```{r}
my_model = lm(y ~ x)
predict(my_model, newdata = data.frame(x = 2050))
```

## Checking the model

- Just finding the best values of the parameters and their uncertainty is not the whole story
- We need to check the fit of the model
- We can do this by analysing whether the assumed probability distribution is correct or not
- Can look at the probability distribution we have fitted to the data, but most people just look at the leftover bits - the residuals

## Residual plot

- Common to creata QQ-plot of the residuals, and a scatter plot of the residuals vs the fits
```{r, fig.height = 4}
par(mfrow=c(1, 2))
qqnorm(my_model$residuals); qqline(my_model$residuals)
plot(my_model$fitted.values, my_model$residuals)
```

## Transforming the data

- Sometimes the residuals of a linear regression look a little bit mis-shapen
- We might improve the fit by adding more covariates, or by transforming the data (the response and/or the covariates)
- If your variables have very large values then you might get better results by standardising your data (subtracting the mean and dividing by the standard deviation)
- Common transformations include the log or square root
- A common transformation in time series data is the _Box-Cox_ transformation...

## Box-Cox

- The Box-Cox transformation is:
$$f(x; \lambda) = \frac{x^\lambda - 1}{\lambda}\mbox{ if }\lambda \ne 0$$
or
$$f(x; \lambda) = \log(x)\mbox{ if }\lambda =0$$
- The usual reason to use it is when the data are _skewed_ and we want it to look more symmetrical
- You need to choose the value of $\lambda$; usually trial and error


## From LMs to GLMs

- If a normal distribution is not suitable for the residuals we need to choose another probability distribution
- Here is some data from an experiment on whitefly:
```{r}
whitefly = read.csv('../../data/whitefly.csv')
head(whitefly, 4)
```
- The `live` and `n` columns indicate how many whitefly survived and were used in the experiment respectively
- We have a fixed total and a number of surviving whitefly out of this total. Which probability distribution might be appropriate?

## Plotting the whitefly data

```{r, fig.height = 4}
barplot(table(whitefly$live))
```
- Doesn't really show the whole story as we're ignoring the total

## Second go

```{r, fig.height = 4}
hist(whitefly$live/whitefly$n, breaks = 30)
```
- Better, but actually there is also a covariate in the number of immature whitefly that were included (variable `imm`)

## Third go

```{r, fig.height = 6}
plot(whitefly$imm, whitefly$live/whitefly$n)
```
- Looks like the more immature whitefly there were, the more likely they were to survive

## Fitting a model

- Let's call $y$ the number of live whitefly at the end of the experiment, and $n$ the number of whitefly used in the experiment
- Let's fit a binomial distribution:
$$y \sim Bin(n, p)$$
- We know the value of $n$ so it is not really a parameter but a fixed part of the data
- We need to estimate $p$
- We could use method of moments or maximum likelihood. If we use method of moments we get $\hat{p}$ = 0.245

## Fitting a better model

- What if we wanted to include `imm` as a covariate?
- One way would be to let $p = \alpha + \beta x$ where $x$ is the number of immature whitefly
- We could fit this using maximum likelihood to get estimates of $\hat{\alpha}$ and $\hat{\beta}$
- This is now a _Generalised Linear Model_ (GLM)
- The likelihood would be:
```{r}
y = whitefly$live
n = whitefly$n
x = whitefly$imm
alpha = 0.5
beta = 0.0001
sum(dbinom(y, n, alpha + beta * x, log = TRUE))
```  
  
## Fitting an even better model
  
- Suppose after using maximum likelihood we got $\hat{\alpha} = 0.2$ and $\hat{\beta} = 0.01$. Then the likelihood would be 
```{r}
alpha = 0.2
beta = 0.01
sum(dbinom(y, n, alpha + beta * x, log = TRUE))
```
- Can anyone see any problems with this model? (Hint: suppose I wanted to predict what proportion would die when $x = 100$)

## Plotting the fitted model

```{r}
plot(x, y/n)
lines(x, alpha + beta * x, col = 'red')
```

## The logit function

\small 
- To stop the lines going out of the range (0, 1) people often use the _logit_ transformation:
$$\log \left[ \frac{p}{1-p} \right] = \alpha + \beta x \mbox{ or } p = \frac{e^{\alpha + \beta x}}{e^{\alpha + \beta x} + 1}$$
- The latter is known as the inverse logit function

- We now maximise the likelihood:
```{r}
p = exp(alpha + beta * x) / (exp(alpha + beta * x) + 1)
sum(dbinom(y, n, p, log = TRUE))
```
- These logit and inverse logit functions are in the boot package:
```{r}
library(boot)
sum(dbinom(y, n, inv.logit(alpha + beta * x), log = TRUE))
```



## Plotting the fit

- Suppose under this method we got maximum likelihood estimates of $\hat{\alpha} = -2$ and $\hat{\beta} = 0.05$
- A plot of the fitted values is now:
```{r, fig.height = 4}
alpha = -2
beta = 0.05
plot(x, y/n)
lines(x, inv.logit(alpha + beta * x), col = 'red')
```

## Finding the maximum likelihood values

- R has a function called `glm` to find the maximum likelihood values for us
- For binomial model with a logit _link function_ we would type:
```{r}
glm(cbind(y, n) ~ x, family = binomial(link = logit))
```

## Further details about the fit

- The interpretation of the $\hat{\beta}$ value (the coefficient of $x$) is in terms of _log odds_. A unit increase in $x$ gives a $\exp(\hat{\beta}) \approx 1.06$ times increase in the probability of a whitefly dying
- You'll see amongst the output something called the _deviance_. This is minus twice the log-likelihood
- It's a common measure used to compare models as the deviance for a linear regression model is just the mean square error
- Next to it you'll see the _Akaike Information Criterion_ or AIC value, which penalises the deviance by adding on twice the number of parameters (i.e. a measure of the complexity of the model)
- Often, you would fit multiple models with different covariates and choose the one(s) with the smallest AIC

## Other glms

- Another common glm is the Poisson, useful for count data
- For example, suppose we treated the number of whitefly who died as a count, and temporarily ignored the $n$ values
- We could fit:
```{r}
glm(y ~ x, family = poisson(link = log))
```
- Recall that the parameter in the Poisson probability distribution represents the mean (and the variance) which must be positive. 
- Like the logit, the log link stops the rate parameter from going negative

## A final word on glms

- There are lots of different types of GLMs We can do Gamma, Negative Binomial, Beta, Inverse Gaussian, ...
- Each has a link function which transforms the main parameter into an unrestricted range through which we can include covariates
- It's also simple to include extra covariates or interactions:
```{r, eval = FALSE}
glm(y ~ x1 + x2 + x1:x2, family = poisson(link = log))
```

- Residual checks are still important, and R will create them for us 
- We can get at them via e.g.
```{r, eval = FALSE}
my_model = glm(cbind(y, n) ~ x, family = binomial(link = logit))
plot(my_model)
```

## Summary 

- Linear Regression and Generalised Linear Models are two common ways to extend standard probability distributions to include covariates
- We estimate the parameters via maximum likelihood using e.g. `lm` or `glm`
- We sometimes need to include a link function which transforms the parameters into an unrestricted range
- There are lots of different types of GLM for every flavour of probability distribution