---
title: 'Class 1: Time Series Models for Ecologists'
author: Andrew Parnell with Emma Howard and John O'Sullivan\newline \texttt{andrew.parnell@ucd.ie} \newline \vspace{1cm}
  \newline \includegraphics[width=1cm]{../UCDlogo.pdf}
output:
  beamer_presentation:
    includes:
      in_header: ../header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'pdf')
par(mar=c(3,3,2,1), mgp=c(2,.7,0), tck=-.01,las=1)
#setwd("~/GitHub/ecots/slides/day_1")
pkgs = c('R2jags','rjags', 'lubridate', 'tidyverse','forecast', 'rstan')
lapply(pkgs, library, character.only = TRUE)
```

## Introductions

- Tell us who you are, what you are working on, and what you hope to get out of the week
- Timetable for the week
- A quick note about pre-requisites

## How this course works

- This course lives on GitHub, at [github.com/andrewcparnell/ecots](www.github.com/andrewcparnell/ecots) which means anyone can see the slides, code, etc, and make comments on it
- The timetable html document provides links to all the pdf slides, handouts, data and practicals
- Let me know if you spot mistakes, as these can be easily updated on the GitHub page
- There is an issues page if you want to ask questions

## R code, slides, and practicals

- All the slides and practicals are available in pdf format for you to annotate
- In the background, the slides and the practicals are written in `Rmarkdown` format, which means you can load them up in Rstudio and see how everything was created
- When you have spare time, feel free to load up the `.Rmd` files and run the code in the background

## R code in slides

- Many of the slides contain R code and output (some of which may be hidden in the `.Rmd` file)
- An example:
```{r, fig.height = 5}
lynx = read.csv(file = '../../data/lynx.csv')
with(lynx, plot(year, number, type = 'l'))
```

## Course format and other details

- Lectures will take place in the morning, practical classes in the afternoon
- Please ask lots of questions
- Some good books:

    - _Forecasting: Principles and Practice_ by Hyndman and Athanasopoulos
    - _Hierarchical Modeling and Inference in Ecology_ by Royle and Dorazio
    - _Bayesian Methods for Ecology_. by McCarthy
    - _Bayesian Data Analysis_ by Gelman et al

- Looking for data? Try: https://datamarket.com/data/list/?q=provider:tsdl 
- (see also sources in Practical 3)

## What is a time series?

- A time series is any set of data where the response variable is measured over time
- There may be other variables included too (covariates)
- Time may be discrete (1, 2, 3, 4, ...) or continuous (1.7, 2.53, 7.12, ...)
- There may be missing values or outliers
- Occasionally there may be more than one response variable (multivariate time series)

## A time series plot: CO2 data

```{r, echo = FALSE}
CO2 = read.csv(file = '../../data/CO2.csv', na.strings = -99.99)
CO2$date = as.Date(paste(CO2$year, CO2$month ,'01', sep = '-'))
with(CO2, plot(date, CO2_ppm, type = 'l', ylab = 'CO2 (parts per million)', xlab = 'Year'), las = 1)
```

## General features of a time series

- Trend: long term behaviour. May be a straight line or something more complicated
- Seasonal: repeated behaviour. May be yearly, monthly, daily, etc. Likely to be dependent on the time resolution
- Error: Leftover uncertainty beyond the trend and seasonal behaviour. May have interesting statistical patterns.

## Writing time series mathematically

- If we write $y_t$ as the value of the response variable at time $t$ then the series can be _decomposed_ as:
$$y_t = \mbox{trend}_t + \mbox{seasonality}_t + \mbox{error}_t$$

- Most time series models concentrate on the error structure
- Time Series analysis is usually harder if you need to identify the seasonality too

## Participation exercise

- I want to you to look at the time series on the following slides and...
- ... identify the trend. Is it linear or non-linear?
- ... identify the seasonality (if any). Can you estimate the frequency?
- ... look at the residual errors after accounting (in your head) for trend and seasonality. Can you spot any patterns or strange observations?

## Data set 1: Wheat production in Canada

```{r, echo = FALSE}
wheat = read.csv('../../data/wheat.csv')
with(wheat, plot(year, wheat, type = 'l',
                 ylab = 'Wheat production (tonnes)',
                 xlab = 'Year'))
```

## Data set 2: Sheep numbers in Asia

```{r, echo = FALSE}
sheep = read.csv('../../data/sheep.csv')
with(sheep, plot(year, sheep, type = 'l',
                 ylab = 'Sheep numbers',
                 xlab = 'Year'))
```

## Data set 3: Lynx trappings in Canada

```{r, echo = FALSE}
lynx = read.csv('../../data/lynx.csv')
with(lynx, plot(year, number, type = 'l',
                ylab = 'Number of Lynx trapped',
                xlab = 'Year'))
```

## Data set 4: Forest fires in Canada

```{r, echo = FALSE}
ff = read.csv('../../data/forest_fires.csv')
with(ff, plot(year, acres, type = 'l',
                ylab = 'Number of acres',
                xlab = 'Year'))
```

## Data set 5: Geese isotopes 

```{r, echo = FALSE}
geese = read.csv('../../data/geese_isotopes.csv')
geese$date = as.Date(paste(geese$year, geese$month ,geese$day, sep = '-'))
with(geese, plot(date, d13CPl, 
                 ylab = 'd13C',
                 xlab = 'Year'))
```

## Data set 5: Geese isotopes again 

```{r, echo = FALSE}
o = order(geese$julianday)
with(geese, plot(julianday[o], d13CPl[o],
                                  ylab = 'd13C',
                 xlab = 'Day of year'))
```

## Data set 6: Oxygen isotopes in ice

```{r, echo = FALSE, fig.height = 5}
ice = read.csv('../../data/GISP2_20yr.csv')
with(ice, plot(Age, Del.18O, type = 'l', ylab = 'd18O'))
```
```{r}
table(diff(ice$Age))
```

## Goals of time series analysis

- Predict future values of the response variable
- Interpolate or smooth the response variable for missing or non-measured times
- Explain which factors are causing the time series to change
- Understand the underlying behaviour of the time series

## Summary 

This course takes a practical approach, and should help you:

- Understand modern time series modelling techniques
- Get and use tools for thinking about and dealing with uncertainty
- Fit time series models, and make predictions
- Understand your time series data, and the process that generates it
