---
title: 'Class 2: Stochastic volatility models and heteroskedasticity '
author: Andrew Parnell \newline \texttt{andrew.parnell@mu.ie}   \newline \vspace{1cm}
  \newline \includegraphics[width=3cm]{../maynooth_uni_logo.jpg}
  \newline \vspace{1cm}
  https://andrewcparnell.github.io/IntroTSA/
  \newline PRESS RECORD 
output:
  beamer_presentation:
    includes:
      in_header: ../header.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'pdf', fig.height = 5)
par(mar=c(3,3,2,1), mgp=c(2,.7,0), tck=-.01,las=1)
pkgs = c('R2jags','rjags', 'lubridate', 'tidyverse','forecast')
lapply(pkgs, library, character.only = TRUE)
```

## Learning outcomes

- Learn how to model changing variance in a time series
- Understand how to fit ARCH, GARCH and SVM models in JAGS
- Know how to compare and plot the output from these models

## General principles of models for changing variance

- So far we have looked at models where the mean changes but the variance is constant:
$$y_t \sim N(\mu_t, \sigma^2)$$
- In this module we look at methods where instead:
$$y_t \sim N(\alpha, \sigma_t^2)$$
- These are:
  
    - Autoregressive Conditional Heteroskedasticity (ARCH) 
    - Generalised Autoregressive Conditional Heteroskedasticity (GARCH) 
    - Stochastic Volatility Models (SVM)

- They follow the same principles as ARIMA, but work on the standard deviations or variances instead of the mean

- `forecast` doesn't include any of these models so we'll use JAGS. There are other R packages to fit these models

## Extension 1: ARCH 

- An ARCH(1) Model has the form:
$$\sigma_t^2 = \gamma_1 + \gamma_2 \epsilon_{t-1}^2$$
where $\epsilon_{t}$ is the residual, just like an MA model
- Note that $\epsilon_t = y_t - \alpha$ so the above can be re-written as:
$$\sigma_t^2 = \gamma_1 + \gamma_2 (y_{t-1} - \alpha)^2$$
- The variance at time $t$ thus depends on the previous value of the series (hence the autoregressive in the name)
- The residual needs to be squared to keep the variance positive. 
- The parameters $\gamma_1$ and $\gamma_2$ also need to be positive, and usually $\gamma_1 \sim U(0,1)$

## JAGS code for ARCH models

\small
```{r}
model_code = '
model
{
  # Likelihood
  for (t in 1:T) {
    y[t] ~ dnorm(alpha, sigma[t]^-2)
  }
  sigma[1] ~ dunif(0, 1)
  for(t in 2:T) {
    sigma[t] <- sqrt(gamma_1 + gamma_2 * pow(y[t-1] - alpha, 2))
  }

  # Priors
  alpha ~ dnorm(0.0, 100^-2)
  gamma_1 ~ dunif(0, 100)
  gamma_2 ~ dunif(0, 100)
}
'
```

## Reminder: forest fires data

```{r, echo = FALSE}
ff = read.csv('../../data/forest_fires.csv')
with(ff, plot(year, acres, type = 'l',
                ylab = 'Number of acres',
                xlab = 'Year'))
```

## ARCH(1) applied to forest fires data

```{r, include=FALSE}
# Run the model - requires longer to converge
ff_run = jags(data = list(y = scale(ff$acres)[,1], T = nrow(ff)),
                     parameters.to.save = c('sigma','alpha','gamma_1','gamma_2'),
                     model.file=textConnection(model_code))
```
```{r}
plot(ff_run)
```

## Plot the ARCH parameters

```{r, fig.height = 6}
par(mfrow=c(1,2))
hist(ff_run$BUGSoutput$sims.list$gamma_1, breaks=30)
hist(ff_run$BUGSoutput$sims.list$gamma_2, breaks=30)
```

## Plot the posterior standard deviations

```{r, echo = FALSE}
par(mfrow=c(2,1))
with(ff, plot(year, acres, type = 'l',
                ylab = 'Number of acres',
                xlab = 'Year',
              main = 'Time Series of forest fires'))
plot(1919:1988, ff_run$BUGSoutput$median$sigma[-1], type = 'l', ylab = 'sigma', main = 'Posterior standard deviation')
sigma_post = ff_run$BUGSoutput$sims.list$sigma
upp = apply(sigma_post, 2, quantile, 0.75)
low = apply(sigma_post, 2, quantile, 0.25)
lines(1919:1988, low[-1], col = 'red', lty = 2)
lines(1919:1988, upp[-1], col = 'red', lty = 2)
```

## From ARCH to GARCH

- The Generalised ARCH model works by simply adding the previous value of the variance, as well as the previous value of the observation
- The GARCH(1,1) model thus has:
$$\sigma_t^2 = \gamma_1 + \gamma_2 (y_{t-1} - \alpha)^2 + \gamma_3 \sigma_{t-1}^2$$
- There are, as always, complicated restrictions on the parameters, though like the stationarity conditions in ARIMA models we can relax this assumption and see if the data support it
- It's conceptually easy to extend to general GARCH(p,q) models which add in extra previous lags

## Example of using the GARCH(1,1) model

\small

```{r}
model_code = '
model
{
  # Likelihood
  for (t in 1:T) {
    y[t] ~ dnorm(alpha, sigma[t]^-2)
  }
  sigma[1] ~ dunif(0,1)
  for(t in 2:T) {
    sigma[t] <- sqrt(gamma_1 + gamma_2 * pow(y[t-1] - alpha, 2) 
                        + gamma_3 * pow(sigma[t-1], 2))
  }
  # Priors
  alpha ~ dnorm(0, 10^-2)
  gamma_1 ~ dunif(0, 10)
  gamma_2 ~ dunif(0, 10)
  gamma_3 ~ dunif(0, 10)
}
'
```

## Using the forest fire data again
```{r, include=FALSE}
# Run the model - requires longer to converge
ff_run_2 = jags(data = list(y = scale(ff$acres)[,1], T = nrow(ff)),
                parameters.to.save = c('sigma', 'alpha', 'gamma_1', 'gamma_2', 'gamma_3'),
                model.file=textConnection(model_code))
```
```{r}
plot(ff_run_2)
```

## Looking at the GARCH parameters

```{r, echo = FALSE}
par(mfrow=c(1,3))
hist(ff_run_2$BUGSoutput$sims.list$gamma_1, breaks=30, xlab = 'gamma 1')
hist(ff_run_2$BUGSoutput$sims.list$gamma_2, breaks=30, xlab = 'gamma 2')
hist(ff_run_2$BUGSoutput$sims.list$gamma_3, breaks=30, xlab = 'gamma 3')
```

## Posterior standard deviations over time

```{r, echo = FALSE}
par(mfrow=c(2,1))
with(ff, plot(year, acres, type = 'l',
                ylab = 'Number of acres',
                xlab = 'Year',
              main = 'Time Series of forest fires'))
plot(1919:1988, ff_run_2$BUGSoutput$median$sigma[-1], type = 'l', ylab = 'sigma', main = 'Posterior standard deviation')
sigma_post = ff_run_2$BUGSoutput$sims.list$sigma
upp = apply(sigma_post, 2, quantile, 0.75)
low = apply(sigma_post, 2, quantile, 0.25)
lines(1919:1988, low[-1], col = 'red', lty = 2)
lines(1919:1988, upp[-1], col = 'red', lty = 2)
```

## Compare with DIC

```{r, echo=FALSE, message=FALSE, results='hide'}
r_1 = print(ff_run)
r_2 = print(ff_run_2)
```
```{r}
with(r_1, print(c(DIC, pD)))
with(r_2, print(c(DIC, pD)))
```

- Suggests not much difference between the models

## Stochastic Volatility Modelling

- Both ARCH and GARCH propose a deterministic relationship for the current variance parameter
- By contrast a Stochastic Volatility Model (SVM) models the variance as its own _stochastic process_
- SVMs, ARCH and GARCH are all closely linked if you go into the bowels of the theory
- The general model structure is often written as:
$$y_t \sim N( \alpha, \exp( h_t ) )$$
$$h_t \sim N( \mu + \phi h_{t-1}, \sigma^2)$$
- You can think of an SVM being like a GLM but with a log link on the variance parameter

## JAGS code for the SVM model

\small

```{r}
model_code = '
model
{
  # Likelihood
  for (t in 1:T) {
    y[t] ~ dnorm(alpha, sigma_h[t]^-2)
    sigma_h[t] <- sqrt(exp(h[t]))
  }
  h[1] <- mu
  for(t in 2:T) {
    h[t] ~ dnorm(mu + phi * h[t-1], sigma^-2)
  }

  # Priors
  alpha ~ dnorm(0, 100^-2)
  mu ~ dnorm(0, 100^-2)
  phi ~ dunif(-1, 1)
  sigma ~ dunif(0,100)
}
'
```

## Example of SVMs and comparison of DIC

```{r, include = FALSE}
# Run the model - requires longer to converge
ff_run_3 = jags(data = list(y = scale(ff$acres)[,1], T = nrow(ff)),
                     parameters.to.save = c("alpha", "mu", "phi", "sigma", "h"),
                     model.file=textConnection(model_code))
```
```{r}
plot(ff_run_3)
```

## Look at all the parameters

```{r, echo=FALSE}
par(mfrow=c(2,2))
hist(ff_run_3$BUGSoutput$sims.list$alpha, breaks=30)
hist(ff_run_3$BUGSoutput$sims.list$mu, breaks=30)
hist(ff_run_3$BUGSoutput$sims.list$phi, breaks=30)
hist(ff_run_3$BUGSoutput$sims.list$sigma, breaks=30)
```

## Plot of $\sqrt{\exp(h)}$

```{r, echo = FALSE}
h_post = ff_run_3$BUGSoutput$sims.list$h
upp = apply(h_post, 2, quantile, 0.75)
low = apply(h_post, 2, quantile, 0.25)
plot(1918:1988, sqrt(exp(ff_run_3$BUGSoutput$median$h)), type = 'l', ylab = 'sqrt(exp(h))', ylim = range(sqrt(exp(c(upp, low)))))
lines(1918:1988, sqrt(exp(low)), col = 'red', lty = 2)
lines(1918:1988, sqrt(exp(upp)), col = 'red', lty = 2)
# h_med = sqrt(exp(apply(ff_run_3$BUGSoutput$sims.list$h,2,'quantile',0.5)/2))
```

## Comparison with previous models
```{r, echo=FALSE, message=FALSE, results='hide'}
r_3 = print(ff_run_3)
```

```{r}
with(r_1, print(c(DIC, pD)))
with(r_2, print(c(DIC, pD)))
with(r_3, print(c(DIC, pD)))
```

Much better fit, despite many extra parameters due to $h$!

## Summary

- We know that ARCH extends the ARIMA idea into the variance using the previous values of the series
- We know that GARCH extends ARCH with previous values of the variance too
- We know that SVMs give the variance its own stochastic process
- We can combine these new models with all the techniques we have previously learnt
